Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-01-09 paper/main.tex
add description of rebalancing procedure to GLMM methods section

In the set of analyses reported in Figure~\ref{fig:predictions}, we assessed
whether estimates of participants' knowledge at the embedding coordinates of
individual quiz questions could be used to reliably predict their ability to
correctly answer those questions. In essence, we treated each question a given
participant answered on a given quiz as a ``lecture'' consisting of a single
timepoint, and used Equation~\ref{eqn:prop} to estimate the participant's
knowledge for its embedding coordinate based on their performance on all
\textit{other} questions they answered on that same quiz (``All questions'';
Fig.~\ref{fig:predictions}, top row). Additionally, for each lecture-related
question (i.e., excluding questions about general physics knowledge), we
computed analogous knowledge estimates based on all other questions the
participant answered on the same quiz about (1) the same lecture as the target
question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows), and (2)
the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions},
bottom rows).

>>>>

In the set of analyses reported in Figure~\ref{fig:predictions}, we assessed
whether estimates of participants' knowledge at the embedding coordinates of
individual quiz questions could be used to reliably predict their ability to
correctly answer those questions. In essence, we treated each question a given
participant answered on a given quiz as a ``lecture'' consisting of a single
timepoint, and used Equation~\ref{eqn:prop} to estimate the participant's
knowledge for its embedding coordinate based on their performance on all
\textit{other} questions they answered on that same quiz (``All questions'';
Fig.~\ref{fig:predictions}, top row). Additionally, for each lecture-related
question (i.e., excluding questions about general physics knowledge), we
computed analogous knowledge estimates based on two different subsets of
questions the participant answered on the same quiz: (1) all \textit{other}
questions about the same lecture as the target question (``Within-lecture'';
Fig.~\ref{fig:predictions}, middle rows), and (2) all questions about the other
of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions},
bottom rows).

In performing these analyses, our null hypothesis is that the knowledge
estimates we compute based on the quiz questions' embedding coordinates do
\textit{not} provide useful information about participants' abilities to answer
those questions---in other words, that there is no meaningful difference (on
average) between the knowledge estimates we compute for questions participants
answered correctly and those they answered incorrectly. Specifically, since we
estimate knowledge for a given embedding coordinate as a weighted
proportion-correct score (where each question's weight reflects its
embedding-space distance from the target coordinate; see Eqn.~\ref{eqn:prop}),
if these weights are uninformative (e.g., randomly distributed), then our
estimates of participants' knowledge should be equivalent (on average) to the
\textit{unweighted} proportion of correctly answered questions used to compute
them. In general, for a given participant and quiz, this expected value (i.e.,
that participant's proportion-correct score on that quiz) is the same for any
coordinate in the embedding space (e.g., any lecture timepoint, quiz question,
etc.). However, in the ``All questions'' and ``Within-lecture'' versions of the
analyses shown in Figure~\ref{fig:predictions}, we estimate each participant's
knowledge for each target question using all \textit{other} questions (or all
\textit{other} questions about the same lecture) they answered on the same quiz.
This introduces a systematic dependency between a participant's success on a
target question and their proportion-correct score on the questions available to
estimate their knowledge for it. For example, suppose a participant correctly
answered $n$ out of $q$ questions on a given quiz. If we hold out a single
\textit{correctly} answered question as the target, the proportion of remaining
questions answered correctly would be $\frac{n - 1}{q - 1}$. Whereas if we hold
out a single \textit{incorrectly} answered question, the proportion of remaining
questions answered correctly would be $\frac{n}{q - 1}$. Thus, the proportion of
correctly answered remaining questions (and therefore the null-hypothesized
value of a knowledge estimate computed from them) is always \textit{lower} for
target questions a participant answered correctly than for those they answered
incorrectly.

To correct for this baseline inverse relationship between a participant's
success on a target question and their estimated knowledge for it, we used a
rebalancing procedure that ensured our knowledge estimates for questions each
participant answered correctly and incorrectly were computed from the
\textit{same} proportion of correctly answered questions. For each target
question on a given participant's quiz, we identified all remaining questions
with the opposite ``correctness'' label (i.e., if the target question was
answered correctly, we identified all remaining incorrectly answered questions,
and vice versa). We then held out each of these opposite-label questions, in
turn, along with the target question, and estimated the participant's knowledge
for the target question using all \textit{other} remaining questions. Since each
of these subsets of remaining questions was constructed by holding out one
correctly answered question and one incorrectly answered question from the
participant's quiz, if the participant correctly answered $n$ out of $q$
questions total, then their proportion-correct score on each subset of questions
used to estimate their knowledge for the target question is $\frac{n-1}{q-2}$,
regardless of whether they answered the target question correctly or
incorrectly. Finally, averaging over these per-subset knowledge estimates
yielded a rebalanced estimate of the participant's knowledge for the target
question that leveraged information from all remaining questions' embedding
coordinates, but whose expected value under our null hypothesis was the same as
that of each individual subset ($\frac{n-1}{q-2}$). By equalizing the
null-hypothesized values of knowledge estimates for correctly and incorrectly
answered questions, this procedure ensures that any meaningful relationships we
observe between participants' estimated knowledge for individual quiz questions
and their ability to answer them are attributable to the predictive power of the
embedding-space distances used to weight questions' contributions to the
knowledge estimates, rather than an artifact of our estimation procedure. Note
that if a participant answered all or no questions on a given quiz correctly,
their responses contained no opposite-label questions with which to perform this
rebalancing, and we therefore excluded their data from our analyses for that
quiz. We used this rebalancing procedure when constructing knowledge estimates
for the ``All questions'' and ``Within-lecture'' versions of the analyses shown
in Figure~\ref{fig:predictions}, but not for the ``Across-lecture'' analyses as,
in this case, the target questions and the questions used to estimate
participants' knowledge for them were drawn from different subsets of quiz
questions (those about one lecture, and those about the other), and were
therefore independent.
